Reagents: expressing and composing fine-grained concurrency

### 2011-11-06 paper/paper.tex
benchmark work, incorporating graphs
writing

\documentclass[preprint]{sigplanconf}

@@

Another answer is: communicate efficiently.  The past two
decades\footnote{Herlihy/Wing, Mellor-Crummey/Scott} have produced a sizable
collection of algorithms for synchronization, communication, and shared
storage which minimize the use of memory bandwidth and avoid unnecessary
waiting.  This research effort has led to industrial-strength
libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering
a wide range of concurrency primitives appropriate for fine-grained
parallelism.

@@

\paragraph{Expressive}  .

\paragraph{Composable} .

We see this work as serving the needs of two distinct groups: concurrency
experts and concurrency users.  Using reagents, experts can write algorithms
more easily, because common patterns are expressible as abstractions and many
are built-in.  Users can compose, extend, and tailor the library without
detailed knowledge of the algorithms involved.

blend shared-state and message-passing concurrency

implemented a Scala library

CLEAR COST MODEL

@@

@@

\cite{Shavit1997a} %TODO: cite pure elimination

@@

Parallelization requires a change of perspective.  Normally, we view concurrent
operations on the same part of a data structure as competing to atomically
update that data structure; we want the operations to be isolated.  On the other
hand, sometimes operations can ``help'' each other: a \lstinline{push} and a
\lstinline{tryPop} effectively cancel each other out.  This insight leads to a
scheme called \emph{elimination backoff}~\cite{?}.  Operations first try the
usual \lstinline{cas}-based code.  If the \lstinline{cas} fails, rather than
busy-waiting, the operations advertise their presence on a side-channel,
reducing the contention for the head pointer.  If a \lstinline{push} and
\lstinline{tryPop} detect their mutual presence, the \lstinline{push} can pass
its argument directly to \lstinline{tryPop} through the side-channel, and no
change to the head pointer is necessary.  Atomicity is not violated, because had
the \lstinline{push} and \lstinline{pop} executed in sequence, the head pointer
would have been returned to its original value anyway.

@@

Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.
They can be invoked using the \lstinline{!} method provided by reagents; for a
\lstinline{Reagent[A,B]} the \lstinline{!} method takes an \lstinline{A} and
returns a \lstinline{B}.  Scala permits infix notation for methods, so we can
use a \lstinline{TreiberStack} as follows:
\begin{lstlisting}
  var s = new TreiberStack[Int]; s.push ! 42
\end{lstlisting}
The key point is that when we invoke these reagents, we are executing
\emph{exactly} the same algorithms written by hand in \secref{background},
including the retry loop with exponential backoff; reagents systematize and
internalize common patterns of fine-grained concurrency.  But by exposing
\lstinline{push} and \lstinline{tryPop} as reagents, rather than methods, we
leave the door open for further composition and tailoring by a user of the
data structure (\secref{choice}, \secref{conjunction}).

@@

\subsection{Synchronization: how reagents react}

@@

@@

@@

@@

bags allow greatest degree of freedom and parallelism

kcas implementations: \cite{Attiya2008,Fraser2007,Luchangco2003}

\begin{lstlisting}
def !(a: A): B = tryReact(a, Reaction.inert, null) match {
  case (_: BacktrackCommand) => {
    val backoff = new Backoff
    val maySync = this.maySync // cache
    @tailrec def retryLoop(shouldBlock: Boolean): B = {
      val wait = maySync || shouldBlock
      val waiter = if (wait) new Waiter[B](shouldBlock) 
                   else null

      tryReact(a, Reaction.inert, waiter) match {
        case (bc: BacktrackCommand) if wait => {
          bc.bottom(waiter, backoff, snoop(a))
          waiter.tryAbort      // rescind waiter,
          waiter.poll match { // but check if completed
            case Some(ans) => ans.asInstanceOf[B] 
            case None       => retryLoop(bc.isBlock)
          }
        }
        case Retry => backoff.once; retryLoop(false)
        case Block => retryLoop(true)
        case ans   => ans.asInstanceOf[B]
      }
    }
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
\end{lstlisting}

\begin{lstlisting}
def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
  r1.tryReact(a, rx, offer) match {
    case Retry => 
      r2.tryReact(a, rx, offer) match {
        case Retry => Retry
        case Block => Retry // must retry r1
        case ans   => ans
      }
    case Block => r2.tryReact(a, rx, offer)
    case ans => ans
  }
\end{lstlisting}

\begin{lstlisting}
  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]): Any = 
    if (rx.canCASImmediate(k, offer)) {
      if (casI(expect, update))
        k.tryReact((), rx, offer)
      else Retry
    } else k.tryReact((), rx.withCAS(Ref.this, expect, update), offer)
\end{lstlisting}

@@

\newpage
\section{Related work}
\label{sec:related}

@@

@@

1.2 Title, abstract, intro,             1.0
1.0 Background                          1.0
2.5 Library overview & examples         3.0
1.5 Semantics 
1.5 Implementation/algorithms           2.0
0.5 Performance results                 1.0
1.0 Related work                        1.0
0.3 Conclusion                          0.3
0.8 Bib                                 0.7

>>>>

\documentclass[preprint,nocopyrightspace]{sigplanconf}

@@

Another answer is: communicate efficiently.  The past two
decades\footnote{Herlihy/Wing~\cite{?}, Mellor-Crummey/Scott~\cite{?}} have
produced a sizable collection of algorithms for synchronization, communication,
and shared storage which minimize the use of memory bandwidth and avoid
unnecessary waiting.  This research effort has led to industrial-strength
libraries---\texttt{java.util.concurrent} (JUC) the most prominent---offering a
wide range of concurrency primitives appropriate for fine-grained parallelism.

@@

\paragraph{Expressive}  
Reagents provide a basic set of building blocks for writing concurrent data
structures and synchronizers.  The building blocks include isolated atomic
updates to shared state, and interactive synchronous communication through
message passing.  This blend of isolation and interaction is essential for
expressing the full range of fine-grained concurrent
algorithms~(\secref{choice}).  The building blocks also bake in many common
concurrency patterns, like optimistic retry loops, backoff schemes, and blocking
and signalling.  Using reagents, it is possible to express sophisticated
concurrency algorithms at a higher level of abstraction, while retaining the
performance and scalability of direct, hand-written implementations.

\paragraph{Composable}
Reagents also provide a set of composition operators, including choice,
sequencing and pairing.  Using these operators, clients can extend and tailor
concurrency primitives without knowledge of the underlying algorithms.  For
example, if a reagent provides only a nonblocking version of an operation like
\lstinline{dequeue}, a user can easily tailor it to a version that blocks if the
queue is empty; this extension will work regardless of how \lstinline{dequeue}
is defined, and will continue to work even if the \lstinline{dequeue}
implementation is changed.  Similarly, clients of reagent libraries can sequence
operations together to provide atomic move operations between arbitrary
concurrent collections---again, without access to or knowledge of the
implementations.  Compositionality is also useful to algorithm designers, since
many sophisticated algorithms can be understood as compositions of simpler
ones~(\secref{choice}, \secref{conjunction}).

\vskip 6pt
\noindent
We begin in~\secref{background} by illustrating, through examples, some of the
common patterns, tradeoffs and concerns in designing fine-grained concurrent
algorithms.  The two subsequent sections contain our core technical
contributions:
\begin{itemize}
\item The \textbf{design} of the reagent combinators is given
  in~\secref{reagents}.  Each combinator is motivated by specific needs and
  patterns in concurrent algorithms; the section shows in
  particular how to write all of the algorithms described in~\secref{background}
  concisely and at a higher-than-usual level of abstraction.
\item The \textbf{implementation} of reagents is detailed
  in~\secref{implementation}, via both high-level discussion and excerpts from
  our code in Scala.  It reveals the extent to which reagents turn patterns of
  fine-grained concurrency into a general algorithmic framework.  It also shows
  how certain choices in the design of reagents enable important optimizations
  in their implementation.
\end{itemize}
Reagents have a clear cost model that provides an important guarantee: they do
not impose extra overhead on the atomic updates performed by individual
concurrent algorithms.  That is, a reagent-based algorithm manipulates shared
memory in exactly the same way as its hand-written counterpart, even if it reads
and writes many disparate locations.  This is a guarantee not generally provided
by concurrency abstractions like software transactional memory (STM), which
employ redo or undo logs or other mechanisms to support composable atomicity.
Reagents only impose overheads when \emph{multiple} algorithms are sequenced
into a large atomic block, \emph{e.g.}  in an atomic transfer between
collections.

\emph{In principle, then, reagents offer a strictly better situation than with
  current libraries}: when used to express the algorithms provided by current
libraries, reagents provide a higher level of abstraction yet impose negligible
overhead; nothing is lost.  But unlike with current libraries, the algorithms
can then be extended, tailored, and combined.  Extra costs are only paid when
the new atomic compositions are used.

We test this principle empirically in~\secref{performance} by comparing multiple
reagent-based collections to their hand-written counterparts, as well as to
lock-based and STM-based implementations.  The benchmarks include both single
operations used in isolation (where little overhead for reagentes is expected)
and combined atomic transfers.  We compare at both high and low levels of
contention.  Reagents perform universally better than the lock- and STM-based
implementations, and are indeed competetive with hand-written lock-free
implementations.

Finally, it is worth noting in passing that reagents generalize both Concurrent
ML~\cite{?} and Transactional Events~\cite{?}, yielding the first lock-free
implementation for both.  We give a thorough discussion of this and other
related work---especially STM---in~\secref{related}.

%CML and TE?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

Parallelization requires a change of perspective.  Normally, we view concurrent
operations on the same part of a data structure as competing to atomically
update that data structure; we want the operations to be isolated.  On the other
hand, sometimes operations can ``help'' each other: a \lstinline{push} and a
\lstinline{tryPop} effectively cancel each other out.  This insight leads to a
scheme called \emph{elimination backoff}~\cite{Shavit1997a,Hendler2004}.
Operations first try the usual \lstinline{cas}-based code.  If the
\lstinline{cas} fails, rather than busy-waiting, the operations advertise their
presence on a side-channel, reducing the contention for the head pointer.  If a
\lstinline{push} and \lstinline{tryPop} detect their mutual presence, the
\lstinline{push} can pass its argument directly to \lstinline{tryPop} through
the side-channel, and no change to the head pointer is necessary.  Atomicity is
not violated, because had the \lstinline{push} and \lstinline{pop} executed in
sequence, the head pointer would have been returned to its original value
anyway.

@@

Being reagents, \lstinline{push} and \lstinline{tryPop} are inert values.  They
can be invoked using the \lstinline{!} method provided by reagents, which is
read ``react''.  For a \lstinline{Reagent[A,B]} the \lstinline{!} method takes
an \lstinline{A} and returns a \lstinline{B}.  Scala permits infix notation for
methods, so we can use a \lstinline{TreiberStack} as follows:
\begin{lstlisting}
  var s = new TreiberStack[Int]; s.push ! 42
\end{lstlisting}
The key point is that when we invoke these reagents, we are executing
\emph{exactly} the same algorithms written by hand in \secref{background},
including the retry loop with exponential backoff; reagents systematize and
internalize common patterns of fine-grained concurrency.  But by exposing
\lstinline{push} and \lstinline{tryPop} as reagents, rather than methods, we
leave the door open for further composition and tailoring by a user of the
data structure (\secref{choice}, \secref{conjunction}).

@@

\subsection{Synchronization: how reagents react}
\label{sec:sync}

@@

% TODO: ``remarkable'' the emergent behavior.  explain in more detail what
% emerges -- the potential for transfer eliminations

@@

%TODO: note tail recursion

@@

Talk about more sophisticated algorithms, e.g. skiplists

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%TODO: talk about dissolve

\elide{
High level concepts:
  Data abstraction/internal normalized syntactic representation
  Continuations/Commit
  Reactions
  Retry versus Block
  Offers

Specific combinators:
  Choice
  CAS - esp. when 1CAS vs kCAS
  Upd?
  Channels - including bag implementation
    bags allow greatest degree of freedom and parallelism
  Computed?
  Sequencing (works by continuation plumbing)

Low-level details:
  kcas implementation
  Inlining
}

Having seen the design and application of reagents, we now turn to their
implementation.  We begin with a high-level overview~(\secref{impl-approach})
introducing the key techniques and data structures we use.  We then delve into
the core code of our Scala implementation, beginning with the primary entry
point (the \lstinline{!} method,~\secref{impl-reaction}) and continuing with the
key combinators.

\subsection{The basic approach}
\label{sec:impl-approach}

When invoked, reagents attempt to \emph{react}, which is \emph{conceptually} a
two phase process: first, the desired reaction is built up; second, the reaction
is atomically committed.  We emphasize ``conceptually'' because, as discussed in
the introduction, reagents are designed to avoid this kind of overhead in the
common case.  We first discuss the general case (which imposes overhead) but
return momentarily to the common (no overhead) case.

An attempt to react can fail during either phase.  A failure during the first
phase, \emph{i.e.} a failure to build up the desired reaction, is always a
permanent failure~(\secref{choice}).  Permanent failures indicate that the
reagent cannot proceed given current conditions, and should therefore block
until another thread intervenes and causes conditions to change.  On the other
hand, a failure during the second phase, \emph{i.e.} a failure to commit, is
always a transient failure~(\secref{choice}).  Transient failures indicate that
the reagent should retry, since the reaction was halted due to active
interference from another thread.  In slogan form, we have
\begin{quotation}\noindent
\begin{tabular}{l@{\ }l}
Permanent failure:&wait for intervention\\
Transient failure:&retry due to interference
\end{tabular}
\end{quotation}
In general, a reaction encompasses three pieces of data:
\begin{enumerate}
\item a list of CASes to be performed,
\item a list of messages to be consumed, and
\item a list of actions to be performed after committing.
\end{enumerate}
It thus resembles the redo log used in some STM implementations~\cite{?}.

In the common case that a reagent only performs one visible~(\secref{dynamic})
CAS or message swap, those components of the reaction are not necessary and
hence are not used.  Instead, the CAS or swap is performed immediately,
compressing the two phases of reaction.  Aside from avoiding extra allocations,
this key optimization means that in the common case a \lstinline{cas} or
\lstinline{upd} in a reagents leads to exactly one executed CAS during reaction,
with no extra overhead.  When a reaction encompasses multiple visible CASes, a
costlier \emph{k}CAS protocol must be used to ensure atomicity.  We discuss the
\emph{k}CAS protocol in~\secref{impl-commit}, and the common case single CAS
in~\secref{impl-state}.

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming afficianados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{BacktrackCommand}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
reagent and the reaction built up so far, and either completes the reaction and
returns its result (type \lstinline{B}) or one of the failure singletons
(\lstinline{Block} or \lstinline{Retry}).

\subsection{The entry point: reacting}
\label{sec:impl-reaction}

shown in \figref{reaction}

\begin{figure}
\begin{lstlisting}
def !(a: A): B = tryReact(a, empty, null) match {
  case (_: BacktrackCommand) => {
    val backoff = new Backoff
    val maySync = this.maySync // cache
    def retryLoop(shouldBlock: Boolean): B = {
      val wait = maySync || shouldBlock
      val waiter = if (wait) new Waiter[B](shouldBlock) 
                   else null

      tryReact(a, empty, waiter) match {
        case (bc: BacktrackCommand) if wait => {
          bc.bottom(waiter, backoff, snoop(a))
          waiter.tryAbort      // rescind waiter,
          waiter.poll match { // but check if completed
            case Some(ans) => ans.asInstanceOf[B] 
            case None       => retryLoop(bc.isBlock)
          }
        }
        case Retry => backoff.once; retryLoop(false)
        case Block => retryLoop(true)
        case ans   => ans.asInstanceOf[B]
      }
    }
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

\subsection{The exit point: committing}
\label{sec:impl-commit}

\subsection{Shared state}
\label{sec:impl-state}

kcas implementations: \cite{Fraser2007,Luchangco2003,Attiya2008}

\begin{lstlisting}
def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
  if (!rx.hasCAS && !k.hasCAS) {
    if (ref.cas(oldV, newV))
      k.tryReact((), rx, offer)
    else Retry
  } else {
    k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
  }
\end{lstlisting}

\subsection{Choice}

\begin{lstlisting}
def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
  r1.tryReact(a, rx, offer) match {
    case Retry => 
      r2.tryReact(a, rx, offer) match {
        case Retry => Retry
        case Block => Retry // must still retry r1
        case ans   => ans
      }
    case Block => r2.tryReact(a, rx, offer)
    case ans => ans
  }
\end{lstlisting}

\subsection{Computed reagents}

\subsection{Channels}

Message passing is by far the most complicated aspect of the reagent
implementation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure*}
\newcommand{\gscale}{0.72}
\hspace{-30pt}
\begin{tabular}[t]{l@{\hspace{2mm}}c}
\begin{sideways}
\hspace{0.5in}
\textbf{Throughput}: \quad
  iterations/$\mu$s \quad
  (bigger is better)
\end{sideways} &
\begin{tabular}[b]{l}
\includegraphics[scale=\gscale]{graphs/PushPop-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/PushPop-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-1000.pdf}
\\
\\
\includegraphics[scale=\gscale]{graphs/EnqDeq-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/EnqDeq-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-1000.pdf}
\end{tabular}
\\
\\
& \textbf{Threads}\quad (on 16-way machine)
\end{tabular}
\end{figure*}

\subsection{Methodology and experimental setup}

Fine-grained concurrent data structures are usually evaluated by targetted
microbenchmarking, with focus on contention effects and fine-grained parallel
speedup~\cite{?}.  In addition to those basic aims, we wish to evaluate (1) the
extent to which reagent-based algorithms can compete with their hand-built
counterparts and (2) whether reagent composition is a plausible approach for
scalable atomic transfers.

To this end, we designed a series of benchmarks

0.25$\mu$s, 2.5$\mu$s

Our benchmarks were run on a 3.46Ghz Intel Xeon X5677 (Westmere) with 32GB RAM
and 12MB of shared L3 cache.  The machine has two physical processors with
four hyperthreaded cores each, for a total of 16 hardware threads.  L1 and L2
caches are per-core.  The software environment includes Ubuntu 10.04.3 and
the Hotspot JVM 6u27.

\subsection{Results}

Performance measurements from our initial implementation show that reagents are
competetive with hand-built concurrent data structures, yet also provide
scalable composed operations that are rarely provided for such data structures.

\subsection{Potential improvements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related work}
\label{sec:related}

@@

We see this work as serving the needs of two distinct groups: concurrency
experts and concurrency users.  Using reagents, experts can write libraries more
easily, because common patterns are expressible as abstractions and many are
built-in.  Users can then extend, tailor and compose the resulting library
without detailed knowledge of the algorithms involved.

@@

1.2 Title, abstract, intro,             1.0
1.0 Background                          1.0
2.5 Library overview & examples         3.0
1.5 Semantics 
1.5 Implementation/algorithms           2.2
0.5 Performance results                 1.0
1.0 Related work                        1.0
0.3 Conclusion                          0.3
0.8 Bib                                 0.5
